---
title: "DDbiplotEZ"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true

bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{DDbiplotEZ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 6, fig.width = 7
)
```

```{r setup}
library(DDbiplotEZ)
```

# Distributional data

The function `create_ddobj` creates an object of class `ddobj` from a dataframe. Included in the package is the dataframe `toy.data`.

```{r, showData}
toy.data
```

This data set consist of 5 different types of data:

* X1 is a numeric variable.
* X2 is an interval scaled variable - the endpoints of the intervals are given in the columns `X2` and `X2.up`. The intention is that the lower bounds appear in the first column and the upper bounds in the second. However, the function `create_ddobj` will reorder the values if they appear in the opposite order, such as for units 4 and 5.
* X3 is a histogram scaled variable - the first group of columns (`X3`, `X3.I1`, `X3.I2`, `X3.I3`, `X3.I4` in this case) contain the consecutive intervals, while the next group of columns (`X3.p1`, `X3.p2`, `X3.p3`, `X3.p4` for the `toy.data`) contain the proportions. If the number of intervals vary between units, the latter columns of units with fewer intervals should contain `NA`. The number of proportions specified, should agree with the number of intervals and the proportions must add up to 1.
* X4 is a categorical (factor in `R`) variable.
* X5 is a modal variable. The format is similar to that of histogram scaled variables where the first group of columns (`X5`, `X5.c2`, `X5.c3` for our example) contains the categories and the latter group the proportions such as `X5.p1`, `X5.p2`, `X5.p3`.

```{r, createData}
out <- create_ddobj(toy.data, type=c("numeric","interval","histogram","categorical", "modal"),
                    cols = c(1, 2, 4, 13, 14), n.int=4, n.cat=3)
```

In addition to specifying the dataframe and type of each variable, the first column for each variable needs to be specified, as well as the number of intervals for histogram scaled variables and the number of categories for modal variables. For multiple histogram scaled or modal variables this number will be a vector.

The print method provides a summary of the `ddobj`.

```{r, showPrint}
print (out)
```

The function `suminto_ddobj` is used to summarise a (usually large) data set into a distributional data object. One such large data set is the `diamonds` data in the package `ggplot2`. 

```{r, getDiamonds, echo=FALSE}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("This vignette requires the 'ggplot2' package.")
}

data("diamonds", package = "ggplot2")
```

We can use the color x cut combinations as units, to summarise carat and price as histogram scaled variables, depth and table as interval scaled variables and clarity as a modal variable.

```{r, showSummarise}
diamonds
suminto_ddobj(diamonds, units = c("color","cut"),
              histogram = c("carat", "price"), interval = c("depth", "table"), modal = "clarity")
```

Two functions allows the user to extract a subset of variables and / or units from a `ddobj`.

```{r, extract}
out
extractDDvars (out, j=c(1,3:4))
extractDDunits (out, i=2:3)
extractDDvars (extractDDunits (out, i=c(2,5)), j=3)
```

## Data sets available in `DDbiplotEZ`

The `Oils.data` from @Ichino1988 is available as well as the `Creditcard.data`. Data sets from the R packages `MAINT.Data` and `dataSDA` can be imported as `ddbiplot` objects using the functions listed below. The function names are exactly those in the package, with a prefix `get.` added.

### Data sets from `MAINT.Data`

* `get.Cars()`
* `get.ChinaTemp()`
* `get.FlightsIdt()`
* `get.LoansbyPurpose_minmaxDt()`
* `get.LoansbyRiskLvs_minmaxDt()`
* `get.LoansbyRiskLvs_qntlDt()`

### Data sets from `dataSDA`

* `get.Abalone()`
* `get.age_cholesterol_weight.int()`
* `get.airline_flights()`
* `get.baseball.int()`
* `get.bird.int()`
* `get.blood_pressure.int()`
* `get.car.int()`
* `get.Face.iGAP()`
* `get.finance.int()`
* `get.horses.int()`
* `get.lackinfo.int()`
* `get.mushroom()`
* `get.profession.int()`
* `get.soccer.bivar.int()`
* `get.veterinary.int()`

### Data sets from `HistDAWass`

* `get.Age_Pyramids_2014()`
* `get.Agronomique()`
* `get.BLOOD()`
* `get.BloodBRITO()`
* `get.China_Month()`
* `get.China_Seas()`
* `get.OzoneFull()`

# Descriptive statistics

First we will construct a small example data set with $n=3$ units, to illustrate the computation of mean, variances and covariances. 

```{r, makeEx_dat}
temp <- cbind (X1 = c(1, 2, 3), X1.b = c(2, 4, 9),
               X2 = c(3, 8, 5), X2.b = c(4, 8.5, 7),
               X3 = c(0, 1, 0), X3.1 = c(0.5, 2, 1), X3.2 = c(1,3,2),
               X3.p1 = c(0.3, 0.4, 0.6), X3.p2 = c(0.7, 0.6, 0.4),
               X4 = c(0, 0, 0), X4.1 = c(1, 1, 1), X4.2 = c(2, 2, 2), X4.3 = c(3, 3, 3),
               X4.p1 = c(0.3, 0.1, 0.7), X4.p2 = c(0.2, 0.6, 0.2), X4.p3 = c(0.5, 0.3, 0.1))
rownames(temp) <- c("A", "B", "C")
ex.dat <- create_ddobj(temp, types = c("interval", "interval", "histogram", "histogram"),
                             cols  = c(    1,          3,           5,           10), n.int=2:3)
ex.dat
```

## Notation

### Interval scaled variables

The following notation will be used:

$$
\mathfrak{X}:n \times p = \begin{bmatrix}
                          \mathfrak{x}_{11} & \dots & \mathfrak{x}_{1p}\\
                          \vdots   &       & \vdots\\
                          \mathfrak{x}_{n1} & \dots & \mathfrak{x}_{np}
                          \end{bmatrix}
                        = \begin{bmatrix}
                          \begin{bmatrix} x^l_{11}, &x^u_{11} \end{bmatrix} & \dots & \begin{bmatrix} x^l_{1p}, & x^u_{1p} \end{bmatrix}\\
                         \vdots   &       & \vdots\\
                          \begin{bmatrix} x^l_{n1}, & x^u_{n1} \end{bmatrix} & \dots & \begin{bmatrix} x^l_{np}, & x^u_{np} \end{bmatrix} 
                          \end{bmatrix}
$$

### Histogram scaled variables

The following notation will be used:

$$
\mathfrak{X}:n \times p = \begin{bmatrix}
                          \mathfrak{x}_{11} & \dots & \mathfrak{x}_{1p}\\
                          \vdots   &       & \vdots\\
                          \mathfrak{x}_{n1} & \dots & \mathfrak{x}_{np}
                          \end{bmatrix}
$$

$$
= \begin{bmatrix}
                          \begin{bmatrix} x^l_{111}, &x^u_{111} \end{bmatrix},p_{111}; 
                          \dots 
                          \begin{bmatrix} x^l_{11s_{11}}, &x^u_{11s_{11}}         \end{bmatrix},p_{11s_{11}} 
                          & \dots & 
                          \begin{bmatrix} x^l_{1p1}, &x^u_{1p1} \end{bmatrix},p_{1p1}; 
                          \dots 
                          \begin{bmatrix} x^l_{1ps_{1p}}, &x^u_{1ps_{1p}}         \end{bmatrix},p_{1ps_{1p}}\\
                         \vdots   &       & \vdots\\
                          \begin{bmatrix} x^l_{n11}, &x^u_{n11} \end{bmatrix},p_{n11}; 
                          \dots 
                          \begin{bmatrix} x^l_{n1s_{11}}, &x^u_{n1s_{n1}}         \end{bmatrix},p_{n1s_{n1}} 
                          & \dots & 
                          \begin{bmatrix} x^l_{np1}, &x^u_{np1} \end{bmatrix},p_{np1};
                          \dots 
                          \begin{bmatrix} x^l_{nps_{n1}}, &x^u_{nps_{n1}}         \end{bmatrix},p_{nps_{np}} 
                          \end{bmatrix}
$$

## Distances

To compute pairwise distances between units, the $l_2$ Wasserstein distance is used, as described in @IrpinoVerde2015. The function `sqL2Wass_dist()` computes the squared $l_2$ Wasserstein distance. It makes used of the function `WassSqDistH()` in the R package `HistDAWass` to compute the squared $l_2$ Wasserstein distance between two units on a histogram scaled variable. Interval scaled variables are converted to histogram scaled variables with a single bin and proportion one, while numeric variables are first converted to interval scaled variables with identical lower and upper interval endpoints, before being converted to histogram scaled variables.

Consider each element of $\mathfrak{X}$, $\mathfrak{X}_{ij}$ to have a density function

$$
\phi_{ij}(x) = \begin{cases}
  \frac{p_{ijh}}{x^u_{ijh}-x^l_{ijh}} & \text{if } x^l_{ijh} < x < x^u_{ijh} \\
  0 & \text{otherwise}
               \end{cases}
$$
with associated distribution function

$$
\Phi_{ij}(x) = \begin{cases}
  0 & x < x^l_{ij1} \\
  \sum_{h=0}^{l-1}{p_{ijh}} + \frac{x - x^l_{ijl}}{x^u_{ijl}-x^l_{ijl}} & l=1,\dots,s_i \text{ if } x^l_{ijl} < x < x^u_{ijl} \\
  1 & x > x^u_{ijs_i}
               \end{cases}
$$

with $p_{ij0} = 0$ and quantile function

$$
\Phi^{-1}_{ij}(t) = x^l_{ij} + (t-\Phi_{ij}(x^l_{ij}))(x^u_{ij} - x^l_{ij})/p_{ijl}, \hspace{1cm} \Phi_{ij}(x^l_{ij}))<t<\Phi_{ij}(x^u_{ij})).
$$

The squared $l_2$ Wasserstein distance between $\phi_{ij}$ and $\phi_{hj}$ with respectively quantile functions $\Phi^{-1}_{ij}$ and $\Phi^{-1}_{hj}$ is computed as

$$
d^2_W(\phi_{ij}, \phi_{hj}) = \int_{0}^{1} [\Phi^{-1}_{ij}(t) - \Phi^{-1}_{hj}(t)]^2\,dt
$$

The squared distances are summed over the $p$ variables, to obtain the squared distance between a pair of units.

```{r, distMatrix}
Dsq <- sqL2Wass_dist(ex.dat)
Dmat <- sqrt(Dsq)
Dmat
```

## Frechet mean 

Following the discussion of @IrpinoVerde2015 the mean distribution of $n$ distributional units, called the Frechet mean, based on the $l_2$ Wasserstein distance is the distribution associated with the quantile function

$$
\bar{\Phi}^{-1}(t) = \frac{1}{n} \sum_{i=1}^n{\Phi^{-1}_i(t)}.
$$

The function `ddQmean` returns this quantile function.

```{r, meanQuantile}
ddQmean (ex.dat, j=1)(seq(from = 0, to = 1, len=21))
ddQmean (ex.dat, j=4)(seq(from = 0, to = 1, len=21))
```

## Variance and covariance based on the Wasserstein distance

@IrpinoVerde2015 propose the following:

$$
s^2_{j} = \frac{1}{n} \sum_{i=1}^{n}{\int_{0}^{1} [\Phi^{-1}_{ij}(t) - \Phi^{-1}_{j}(t)]^2\,dt}
$$

and covariance

$$
s(\mathfrak{X}_j, \mathfrak{X}_k) = \frac{1}{n} \sum_{i=1}^{n}{\int_{0}^{1} [\Phi^{-1}_{ij}(t) - \Phi^{-1}_{j}(t)] [\Phi^{-1}_{ik}(t) - \Phi^{-1}_{k}(t)]\,dt}.
$$

### The function `ddvarW()`

The function accepts an object of class `ddobj` and computes a variance-covariance matrix. 

```{r, Wasserstein_var}
ddvarW(ex.dat)
```

### The function `ddcorW()`

The function uses the variance-covariance matrix produced by `ddvarW()` to divide each covariance by the product of the relevant standard deviations to produce a correlation matrix.

```{r, Wasserstein_cor}
ddcorW(ex.dat)
```

# Descriptive statistics (Billard, 2008)

These computations for mean, variance and covariance based on @Billard2008 can also be computed, although it is not used in the biplot computations.

## Interval scaled variables

### Sample mean

$$
\bar{x}_j = \frac{1}{2n} \sum_{i=1}^{n}(x^u_{ij} + x^l_{ij})
$$

As an example, we can compute the sample mean of variable $X_1$.

$$
\begin{bmatrix}
\begin{bmatrix} 1, & 2 \end{bmatrix}\\
\begin{bmatrix} 2, & 4 \end{bmatrix}\\
\begin{bmatrix} 3, & 9 \end{bmatrix} 
\end{bmatrix}
$$

$$
\bar{x}_1 = \frac{(2+1) + (4+2) + (9+3)}{2 \times 3} = 3.5
$$
```{r, X1_mean}
ddmean(ex.dat$X1)
```

### Sample variance

$$
s_j^2 = \frac{1}{3n} \sum_{i=1}^{n}({x^u_{ij}}^2 + x^u_{ij} x^l_{ij} + {x^l_{ij}}^2) - \bar{x}_j^2
$$

In the example below, we compute the sample variance of variable $X_2$.

$$
\begin{bmatrix}
\begin{bmatrix} 3, & 4 \end{bmatrix}\\
\begin{bmatrix} 8, & 8.5 \end{bmatrix}\\
\begin{bmatrix} 5, & 7 \end{bmatrix} 
\end{bmatrix}
$$
$$
\bar{x}_2 = 5.916667
$$
$$
s_2^2 = \frac{(4^2 + (4)(3) + 3^2) + (8.5^2 + (8.5)(8) + 8^2) + (7^2 + (7)(5) + 5^2)}{3 \times 3} - (5.916667)^2
$$

$$
= \frac{37 + 204.25 + 109}{9} - 35.0070 = 3.9097
$$


```{r, X2_var, error = FALSE}
ddvarB(ex.dat$X2)
```

### Sample covariance

$$
cov(\mathfrak{X}_j,\mathfrak{X}_k) = \frac{1}{6n} \sum_{i=1}^{n}\left[2(x^l_{ij} - \bar{x}_j)(x^l_{ik} - \bar{x}_k) + (x^l_{ij} - \bar{x}_j)(x^u_{ik} - \bar{x}_k) + (x^u_{ij} - \bar{x}_j)(x^l_{ik} - \bar{x}_k) + 2(x^u_{ij} - \bar{x}_j)(x^u_{ik} - \bar{x}_k) \right] 
$$

The covariance of $X_1$ and $X_2$ is computed below.

$$
\begin{bmatrix}
\begin{bmatrix} 1 & 2 \end{bmatrix} & \begin{bmatrix} 3 & 4 \end{bmatrix} \\
\begin{bmatrix} 2 & 4 \end{bmatrix} & \begin{bmatrix} 8 & 8.5 \end{bmatrix} \\
\begin{bmatrix} 3 & 9 \end{bmatrix} & \begin{bmatrix} 5 & 7 \end{bmatrix} 
\end{bmatrix}
$$

$$
\bar{x}_1 = 3.5, \hspace{1cm} \bar{x}_2 = 5.916667
$$
$$
cov(\mathfrak{X}_1,\mathfrak{X}_2) = \frac{\left[2(x^l_{11} - \bar{x}_1)(x^l_{12} - \bar{x}_2) + (x^l_{11} - \bar{x}_1)(x^u_{12} - \bar{x}_2) + (x^u_{11} - \bar{x}_1)(x^l_{12} - \bar{x}_2) + 2(x^u_{11} - \bar{x}_1)(x^u_{12} - \bar{x}_2) \right]}{6 \times 3}  + 
$$
$$
\frac{\left[2(x^l_{21} - \bar{x}_1)(x^l_{22} - \bar{x}_2) + (x^l_{21} - \bar{x}_1)(x^u_{22} - \bar{x}_2) + (x^u_{21} - \bar{x}_1)(x^l_{22} - \bar{x}_2) + 2(x^u_{21} - \bar{x}_1)(x^u_{22} - \bar{x}_2) \right]}{6 \times 3}  +
$$
$$
\frac{\left[2(x^l_{31} - \bar{x}_1)(x^l_{32} - \bar{x}_2) + (x^l_{31} - \bar{x}_1)(x^u_{32} - \bar{x}_2) + (x^u_{31} - \bar{x}_1)(x^l_{32} - \bar{x}_2) + 2(x^u_{31} - \bar{x}_1)(x^u_{32} - \bar{x}_2) \right]}{6 \times 3}
$$

$$
 = \frac{\left[2(1 - 3.5)(3 - 5.92) + (1 - 3.5)(4 - 5.92) + (2 - 3.5)(3 - 5.92) + 2(2 - 3.5)(4 - 5.92) \right]}{18}  + 
$$

$$
\frac{\left[2(2 - 3.5)(8 - 5.92) + (2 - 3.5)(8.5 - 5.92) + (4 - 3.5)(8 - 5.92) + 2(4 - 3.5)(8.5 - 5.92) \right]}{18}  +
$$

$$
\frac{\left[2(3 - 3.5)(5 - 5.92) + (3 - 3.5)(7 - 5.92) + (9 - 3.5)(5 - 5.92) + 2(9 - 3.5)(7 - 5.9) \right]}{18}
$$

$$
= \frac{29.5 - 6.5 + 7.25}{18} = 1.68
$$

```{r, X1_X2_cov, error = FALSE}
ddcovB(ex.dat$X1, ex.dat$X2)
ddvarB(ex.dat$X1, ex.dat$X2)
```

Calling the `varB()` function with two arguments is equivalent to calling the function `covB()`.

## Histogram scaled variables

### Sample mean

$$
\bar{x}_j = \frac{1}{2n} \sum_{i=1}^{n}\left[ \sum_{h=1}^{s_h}{(x^u_{ijh} + x^l_{ijh})p_{ijh}} \right]
$$

where $a_{ih}$ is the lower endpoint of the $h$th bin and $b_{ih}$ is the upper endpoint of the $h$th bin for the histogram of unit $i$. The proportion of observations in bin $[a_{ih},b_{ih}]$ is $p_{ih}$ with a total of $s_h$ bins for the histogram of the variable.

As an example, we can compute the sample mean of variable $X_4$.

$$
\begin{bmatrix}
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.3; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.2; &   \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.5\\
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.1; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.6, & \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.3\\
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.7; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.2; & \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.1
\end{bmatrix}
$$

$$
\bar{x}_4= \frac{(0+1)0.3 + (1+2)0.2 + (2+3)0.5}{2 \times 3} + \frac{(0+1)0.1 + (1+2)0.6 + (2+3)0.3}{2 \times 3}
$$

$$
 + \frac{(0+1)0.7 + (1+2)0.2 + (2+3)0.1}{2 \times 3}
$$

$$
= \frac{0.3+0.6+2.5}{6} + \frac{0.1+1.8+1.5}{6} + \frac{0.7+0.6+0.5}{6} = 1.433
$$

```{r, X4_mean}
ddmean(ex.dat$X4)
```

### Sample variance

$$
s_j^2 = \frac{1}{3n} \sum_{i=1}^{n}\left[ \sum_{h=1}^{s_h}{({x^u_{ijh}}^2 + x^u_{ijh}x^l_{ijh} + {x^l_{ijh}}^2)p_{ijh}} \right] - \bar{x}_j^2
$$

We can compute the sample variance of variable $X_3$ with sample mean $\bar{x}_3 = 1.2$.

$$
\begin{bmatrix}
\begin{bmatrix} 0, & 0.5 \end{bmatrix}, 0.3; & \begin{bmatrix} 0.5, & 1 \end{bmatrix}, 0.7\\
\begin{bmatrix} 1, & 2 \end{bmatrix}, 0.4; & \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.6\\
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.6; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.4; 
\end{bmatrix}
$$

$$
s_j^2= \frac{(0+0+0.25)0.3 + (0.25+0.5+1)0.7}{3 \times 3} + \frac{(1+2+4)0.4 + (4+6+9)0.6}{3 \times 3} 
$$

$$
+ \frac{(0+0+1)0.6 + (1+2+4)0.4}{3 \times 3} - 1.2^2
$$

$$
= \frac{0.075+1.225}{9} + \frac{2.8+11.4}{9} + \frac{0.6+2.8}{9} - 1.44 = 0.66
$$

```{r, X3_var}
ddvarB(ex.dat$X3)
```

### Sample covariance

$$
cov(\mathfrak{X}_j,\mathfrak{X}_k) = \frac{1}{6n} \sum_{i=1}^{n}
\left[ \sum_{h=1}^{s_h}\sum_{l=1}^{s_l}{[2(x^l_{ijh} - \bar{x}_j)(x^l_{ikl} - \bar{x}_k) + (x^l_{ijh} - \bar{x}_j)(x^u_{ikl} - \bar{x}_k) \\ + (x^u_{ijh} - \bar{x}_j)(x^l_{ikl} - \bar{x}_k) + 2(x^u_{ijh} - \bar{x}_j)(x^u_{ikl} - \bar{x}_k)]p_{ijh}p_{ikl}} 
\right] 
$$

The sample covariance of variables $X_3$ and $X_4$ is computed below.

$$
\begin{bmatrix}
\begin{bmatrix} 0, & 0.5 \end{bmatrix}, 0.3; & \begin{bmatrix} 0.5, & 1 \end{bmatrix}, 0.7\\
\begin{bmatrix} 1, & 2 \end{bmatrix}, 0.4; & \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.6\\
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.6; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.4; 
\end{bmatrix}
\&
\begin{bmatrix}
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.3; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.2; &   \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.5\\
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.1; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.6, & \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.3\\
\begin{bmatrix} 0, & 1 \end{bmatrix}, 0.7; & \begin{bmatrix} 1, & 2 \end{bmatrix}, 0.2; & \begin{bmatrix} 2, & 3 \end{bmatrix}, 0.1, 
\end{bmatrix}
$$

$$
cov(\mathfrak{X}_3,\mathfrak{X}_4) = \frac{1}{6n} \sum_{i=1}^{n}
\left[ \sum_{h=1}^{2}\sum_{l=1}^{3}{[2(x^l_{i3h} - \bar{x}_3)(x^l_{i4l} - \bar{x}_4) ]p_{i3h}p_{i4l}}
\right] 
$$

$$
+ \frac{1}{6n} \sum_{i=1}^{n}
\left[ \sum_{h=1}^{2}\sum_{l=1}^{3}{[(x^l_{i3h} - \bar{x}_3)(x^u_{i4l} - \bar{x}_4) ]p_{i3h}p_{i4l}} 
\right] 
$$

$$
+ \frac{1}{6n} \sum_{i=1}^{n}
\left[ \sum_{h=1}^{2}\sum_{l=1}^{3}{[(x^u_{i3h} - \bar{x}_3)(x^l_{i4l} - \bar{x}_4) ]p_{i3h}p_{i4l}} 
\right] 
$$

$$
+ \frac{1}{6n} \sum_{i=1}^{n}
\left[ \sum_{h=1}^{2}\sum_{l=1}^{3}{[2(x^u_{i3h} - \bar{x}_3)(x^u_{i4l} - \bar{x}_4)]p_{i3h}p_{i4l}} 
\right] 
$$

Writing the summations in each of the 4 terms, in general as

$$
\sum_{i-1}^{n}\sum_{h=1}^{2}\sum_{l=1}^{3}{(x_{i3h} - \bar{x}_3)(x_{i4l} - \bar{x}_4)p_{i3h}p_{i4l}}
$$

it equals the sum of the elements of the matrix $\mathbf{A'B}$ where

$$
\mathbf{A} = \begin{bmatrix} (x_{131} - \bar{x}_3)p_{131} & (x_{132} - \bar{x}_3)p_{132} \\
                             \vdots                       & \vdots\\
                             (x_{n31} - \bar{x}_3)p_{n31} & (x_{n32} - \bar{x}_3)p_{n32}
             \end{bmatrix}
$$

$$
\mathbf{B} = \begin{bmatrix} (x_{141} - \bar{x}_4)p_{141} & (x_{142} - \bar{x}_4)p_{142} & (x_{143} - \bar{x}_4)p_{143}\\
                             \vdots                       & \vdots\\
                             (x_{n41} - \bar{x}_4)p_{n41} & (x_{n42} - \bar{x}_4)p_{n42} & (x_{n43} - \bar{x}_4)p_{n43}
             \end{bmatrix}.
$$

$cov(\mathfrak{X}_3,\mathfrak{X}_4) =$ sum of

$$
\frac{2}{6 \times 3}
\begin{bmatrix}
(0-1.2)0.3 & (1-1.2)0.4 & (0-1.2)0.6 \\
(0.5-1.2)0.7 & (2-1.2)0.6 & (1-1.2)0.4
\end{bmatrix}
\begin{bmatrix}
(0-1.433)0.3 & (1-1.433)0.2 & (2-1.433)0.5 \\
(0-1.433)0.1 & (1-1.433)0.6 & (2-1.433)0.3 \\
(0-1.433)0.7 & (1-1.433)0.2 & (2-1.433)0.1
\end{bmatrix}
$$

$$
+ \frac{1}{6 \times 3}
\begin{bmatrix}
(0-1.2)0.3 & (1-1.2)0.4 & (0-1.2)0.6 \\
(0.5-1.2)0.7 & (2-1.2)0.6 & (1-1.2)0.4
\end{bmatrix}
\begin{bmatrix}
(1-1.433)0.3 & (2-1.433)0.2 & (3-1.433)0.5 \\
(1-1.433)0.1 & (2-1.433)0.6 & (3-1.433)0.3 \\
(1-1.433)0.7 & (2-1.433)0.2 & (3-1.433)0.1
\end{bmatrix}
$$

$$
+ \frac{1}{6 \times 3}
\begin{bmatrix}
(0.5-1.2)0.3 & (2-1.2)0.4 & (1-1.2)0.6 \\
(1-1.2)0.7 & (3-1.2)0.6 & (2-1.2)0.4
\end{bmatrix}
\begin{bmatrix}
(0-1.433)0.3 & (1-1.433)0.2 & (2-1.433)0.5 \\
(0-1.433)0.1 & (1-1.433)0.6 & (2-1.433)0.3 \\
(0-1.433)0.7 & (1-1.433)0.2 & (2-1.433)0.1
\end{bmatrix}
$$

$$
+ \frac{2}{6 \times 3}
\begin{bmatrix}
(0.5-1.2)0.3 & (2-1.2)0.4 & (1-1.2)0.6 \\
(1-1.2)0.7 & (3-1.2)0.6 & (2-1.2)0.4
\end{bmatrix}
\begin{bmatrix}
(1-1.433)0.3 & (2-1.433)0.2 & (3-1.433)0.5 \\
(1-1.433)0.1 & (2-1.433)0.6 & (3-1.433)0.3 \\
(1-1.433)0.7 & (2-1.433)0.2 & (3-1.433)0.1
\end{bmatrix}
$$

$=$ sum of

$$
\frac{1}{9}
\begin{bmatrix}
-0.36 & -0.08 & -0.72 \\
-0.49 & 0.48 & -0.08
\end{bmatrix}
\begin{bmatrix}
-0.43 & -0.0867 & 0.2835 \\
-0.1433 & -0.26 & 0.1701 \\
-1.0031 & -0.0867 & 0.0567
\end{bmatrix}
$$

$$
+ \frac{1}{18}
\begin{bmatrix}
-0.36 & -0.08 & -0.72 \\
-0.49 & 0.48 & -0.08
\end{bmatrix}
\begin{bmatrix}
-0.13 & 0.1134 & 0.7835 \\
-0.0433 & 0.3402 & 0.4701 \\
-0.3031 & 0.1134 & 0.1567
\end{bmatrix}
$$

$$
+ \frac{1}{18}
\begin{bmatrix}
-0.21 & 0.32 & -0.12 \\
-0.14 & 1.08 & 0.32
\end{bmatrix}
\begin{bmatrix}
-0.43 & -0.0867 & 0.2835 \\
-0.1433 & -0.26 & 0.1701 \\
-1.0031 & -0.0867 & 0.0567
\end{bmatrix}
$$

$$
+ \frac{1}{9}
\begin{bmatrix}
-0.21 & 0.32 & -0.12 \\
-0.14 & 1.08 & 0.32
\end{bmatrix}
\begin{bmatrix}
-0.13 & 0.1134 & 0.7835 \\
-0.0433 & 0.3402 & 0.4701 \\
-0.3031 & 0.1134 & 0.1567
\end{bmatrix}
$$

$=$ sum of

$$
\frac{1}{9}
\begin{bmatrix}
0.8885 & 0.1143 & -0.1565 \\
0.2222 & -0.0754 & -0.0618
\end{bmatrix}
+
\frac{1}{18}
\begin{bmatrix}
0.2685 & -0.1497 & -0.4325 \\
0.0672 & 0.0987 & -0.1708
\end{bmatrix}
$$

$$
+ \frac{1}{18}
\begin{bmatrix}
0.1648 & -0.0546 & -0.0119 \\
-0.4156 & -0.2964 & 0.1622
\end{bmatrix}
+
\frac{1}{9}
\begin{bmatrix}
0.0498 & 0.0714 & -0.0329 \\
-0.1256 & 0.3878 & 0.4482
\end{bmatrix}
$$

$$
= \frac{0.9313}{9} + \frac{-0.3189}{18} + \frac{-0.4515}{18} + \frac{0.7988}{9} = 0.1494
$$

```{r, X3_X4_cov}
ddvarB(ex.dat$X3, ex.dat$X4)
```

The function will also compute the covariance between an interval scaled variable and a histogram scaled variable, by viewing the interval scaled variable as a histogram with a single bin and proportion one.

```{r, X1_X4_cov}
ddcovB(ex.dat$X1, ex.dat$X4)
```

Unlike the function `var()` which accepts a matrix of variables to compute a covariance matrix, the `varB()` function only computes a single numeric variance or covariance value. The function `covmatB()` computes a covariance matrix.

```{r, covarianceMatrix}
ddcovmatB(ex.dat)
```

### Correlation

Once the covariance and variance for interval scaled and histogram scaled variables have been defined, the correlation is computed as usual $cor(x,y) = \frac{cov(x,y)}{s_xs_y}$.

```{r, correlationMatrix}
ddcorB (ex.dat$X2, ex.dat$X3)
ddcormatB(ex.dat)
```

# The function `ddbiplot()`

This function creates an object of class `ddbiplot`. The input is a `ddobj` with a series of optional arguments for grouping units, centering and scaling as well as providing a title for the plot.

The function prepares the `ddobj` for creating biplots. The numeric part with `numeric`, `interval` and `histogram` variables are returned in a component `X`. If there are missing observations, these are removed unit-wise with a warning to the user while the numbers of the units removed $(1, \dots, n)$ is returned in the `na.action` component.

# PCA biplots

The Oils data set (@Ichino1988) on $n=8$ units for 4 interval scaled and 1 numeric variable is included in the `DDbiplotEZ` package. 

```{r, showOils}
Oils.data
```

When the function `PCA()` is called, the numeric variable is converted to an interval scaled variable with $x^l_i = x^u_i$ so that $p=5$ variables are represented.

```{r, PCA_Oils}
ddbiplot (Oils.data, scaled = "W") |>
  PCA () |> plot ()
```

The BLOOD data set (@BillardDiday2006) can be obtained from the `HistDAWass` package with the function `get.BLOOD()`. It contains $n=14$ units and $p=3$ histogram scaled variables.

```{r, showBLOOD}
BLOOD.data <- get.BLOOD()
BLOOD.data
```

```{r, PCA_BLOOD}
blood.data <- get.BLOOD()
ddbiplot (blood.data, scaled = "W") |>
  PCA () |> plot ()
```

### Aesthetics of the units in the PCA biplot

The function `units()` allows the user to choose the aesthetics for the units in the PCA biplot. The following options are available:

* `which` of the groups defined by `classes` or `group.aes` in `PCA()` to show in the biplot.
* `col` the colour for representing units.
* `label` whether the units should be labelled.
* `label.name` names for the units.
* `label.col` colour for unit labels.
* `label.cex` character expansion for unit labels.
* `label.side` one of `bottom`, `left`, `top` or `right` for placing the unit labels.
* `label.offset` how far away the units labels should be positioned.

The argument `col` is typically a vector with length equal to the number of groups specified. The argument `label` is a single `TRUE` or `FALSE` while the other arguments related to the labels are typically vectors of length $n$. For all these vectors recycling is used to ensure the correct length.

```{r, PCAint_aesthetics}
ddbiplot (Oils.data, scaled = "W") |>
  units (label = TRUE, 
         label.side = c("bottom","bottom","right","top","bottom","bottom",
                        "bottom","top")) |>
  PCA () |> plot ()

ddbiplot (blood.data, scaled = "W", group.aes=1:14) |>
  units (col = c("red","blue","darkgreen","purple","orange","gold","tan4",
                 "magenta","grey","cyan","coral","springgreen3","lightslateblue",
                 "mediumvioletred")) |>
  PCA () |> plot ()
```

### More aesthetics from package `biplotEZ`

The aesthetics of the biplot axes are set with the function `axes()` from `biplotEZ`. Similarly a legend is added with the `legend.type(samples = TRUE)` function.

```{r, PCAint_more_aes, fig.height=6, fig.width=7}
ddbiplot (blood.data, scaled = "W", group.aes=1:14) |>
  units (col = c("red","blue","darkgreen","purple","orange","gold","tan4",
                 "magenta","grey","cyan","coral","springgreen3","lightslateblue",
                 "mediumvioletred")) |>
  PCA () |> 
  legend.type (samples = TRUE, new = TRUE) |>
  plot ()
```



# References
